Changes in Mobility and Site Occupation during the Late Pleistocene in Korea

### 2020-01-30 analysis/paper/paper.Rmd
edited by HBE part

<!--do we need to talk about AT???People had assumed that Japanese archipelago started to be occupied before 30,000 cal BP based on the existence of primary Aira-Tanzawa tephra (dated at 30,009±189 cal BP) in the excavation unit. Recenly, accummulated radiocarbon data show that --> 
<<!-- dp we need to mention popular research topic such as existance of pre UP ("archaic industries")as a heated debate?--->>

The Late Pleistocene technological transition in Korea is mainly characterized by the early appearance of stemmed points, followed by blade technology. The stemmed point is a projectile point made out of elongated flake or blade and with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. The earliest stemmped points were made on flakes. After blades appear at c. 27 ka, the shape of stemmped points became more standardized with one or two ridges on dorsal side and triangular cross section. The stemmed point is often called as a tanged point in western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points in Korea that have long been called 'tanged points'. Stemmed points were made from blade blanks, and became more standardized and increased in quantity over time. Flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene  [@Bae_2010; @Lee_2013; @Lee_2016]. The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; #### Park 2012 ####; @seong2015diversity; @bae2017origin]. The origin of the stemmed point is one of the most frequently devated topic related to the modern human dispersals. Bae (2010) claims two sub-stages for these new technologies based on the presence of AT tephra (presumably earlier than 25k BP). Bae’s earlier stage is started from Koreyri site dated to 27 ka. The stage is characterized by long and thick blades and relatively large prismatic cores with stemmed points. Bae’s later stage contains a microlithic industry, for example as found at Sinbuk in southwestern Korea, which dates 25,420±190 BP and 25,000±1,000 BP [@bae2010origin]. An alternative view comes from Seong (2009) and Park (2012) who claim three sub-stages by distinguishing an initial period when stemmed points appeared before blades [@seong2009emergence; #### Park 2012 ####]. The difference between the two stages of division [@bae2010origin] and the three stages [@seong2009emergence; #### Park 2012 ####] hinges on the recognition of stemmed points as a technology that is distinctive from blades. Bae stresses the combination of stemmed point and blade as tool blanks, while Seong and Park focus on the stemmed point as a new technology, having a stem part with retouch and acute-angled tip. These different perspectives on the distinctiveness of the stemmed point are at the crux of a heated debate on the nature and cause of the transition from the earlier phase to the later phase during the Late Pleistocene.

@@

Our review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focussed on timing, location and description of these new tools, but there has been little effort to model the land-use behaviours that generated these assemblages. We focus on the appearance of stemmed points in Korea because these are one of the earliest new technologies that definiately represents a new hunting methods because the halfting implies throwing techniques that can reach long-distance targets [@Keeley_1982; @kuhn2015artifacts]. We ask two questions to invetigate the period around the introduction of stemmed points in Korea.

First, we ask how the introduction of the new tools changed forager landscape use, specifically, was the use of stemmed points associated with occupation of marginal or exterme environments? The new technology might allow foragers to explore and stay in different or harsher territories by increasing mobility. <!-- need something about how technology and landscape use is linked ... - the way people used the *landscape*: changing mobility patterns due to risk management,  Kelly, Fitzhugh, Shott  ... -->

Second, we focus on how the new technology changed the way people use their habitation sites. After the appearance of these new tools, did they tend to stay in one location longer, or for shorter periods, perhaps for specific purposes?. Archaeological sites indicate depositional phenomena and places where artifacts and other debris accumulated reflecting human activities in the past [@kuhn2015artifact]. The stone artifact assemblage from a site can inform us how long individuals or groups stayed and give insignts into what they did there [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface]. Stone artifacts can contain a record of raw material transport and selection and give insights into land-use behaviour [@clarkson2008lithics]. <!-- - the way people used *sites*: more fine-grained evidence about land use patterns long vs short term durations of site use Kuhn, Andresfsky, Hiscock -->

@@

<!--  -------------- previous paragraphs ------

New technologies have led people to adjust the speed and frequency of mobility, explore the new landscape, or even manage the population size of the foraging group or overall population occupied one region. Previous studies show that new technologies play an important role in the change in the use of the landscape. For instance, Morisaki et al. argue that the transition of the blades and points becoming dominant over trapezoid in toolkits around 25 ka cal BP in north Paleo-Honshu Island enabled the foragers to extend their occupation at residential sites [@Morisaki_Izuho_Terry_Sato_2015]. In Australia during the mid-Holocene, the highly retouched and finely shaped tools with shafts appeared to reduce risks of being in a different environment so that they could broaden their occupations to offshore islands on the northeast and northwest coast [@Hiscock_1994]. New tools tend to appear as a risk reduction strategy when they confront an unfamiliar environment mainly caused by climate change, migration to previously unoccupied landscapes, based on the understanding of resource distribution, tool manufacturing process, and the characteristics of the foraging group. The appearance of microliths as portable and standardized tools is a great example to show that people made new tools to increase multi-functionality and tool readiness so that they can minimize the economic risk and/or scheduling difficulties [@Hiscock_Clarkson_Mackay_2011].

The rapid spread of microblade industry in north China among Last Glacial Maximum demonstrates that making new tools was one of the human adaptation to extremely cold environment by increasing mobility or meeting the new demands such as the production of sophisticated cold-weather clothing [@Barton_Brantingham_Ji_2007; @yi2013microblade; @Qu_Bar-Yosef_Wang_Wu_2013; @nian2014chronology].

The explanation of the emergence of new tools as an aid of buffering risks, as a result of environmental change, is one of the main concepts of Human Behavioral Ecology (HBE). HBE focuses on understanding patterned human behaviors, such as the ways they evaluate choices in food procurement, mobility, division of labor, task-group membership, social affiliation, prestige, and social subordination, to selective pressures in the environment [@smith2017natural; @smith2017three; @fitzhugh2001risk; @bird2006behavioral; @fitzhugh2019human; @Veile_2018]. Thus, most HBE models investigate environmental variables (e.g., resource density, competitor frequency) and behaviors reflecting decisions that people made about how to allocate time and energy to their activities [@smith2017three; @bird2006behavioral; @kelly1995foraging; @Creanza_Kolodny_Feldman_2017; @fitzhugh2019human]. HBE provides concepts to identify how foragers adjust their technologies, resource management strategies, schedule movements, foraging, food processing (including storage and consumption), and investment in technologies to optimize time and energy [@kelly1995foraging]. Stone artifacts have also been used in HBE models to understand hunter-gatherers’ mobility and site occupation patterns [@Marwick_2013; @Clarkson_2013; @capriles2018mobility]. For example, Clark and Barton (2017) examined to investigate resource acquisition strategies of hunter-gatherers by means of stone artifact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) as proxies to represent occupation patterns such as ‘logistically organized basecamp’ and ‘short-term camp’. We use HBE as a reference framework to answer the three questions: (1) how did the introduction of the stemmed points and blade industry change the way people used the landscape? (2) was there any change in the way people used sites? (3) was there any population change and did it impact on technology or foraging pattern?

-->

>>>>

<!--do we need to talk about AT???People had assumed that Japanese archipelago started to be occupied before 30,000 cal BP based on the existence of primary Aira-Tanzawa tephra (dated at 30,009±189 cal BP) in the excavation unit. Recenly, accummulated radiocarbon data show that --> 
<<!-- do we need to mention popular research topic such as existance of pre UP ("archaic industries")as a heated debate?--->

The Late Pleistocene technological transition in Korea is mainly characterized by the early appearance of stemmed points, followed by blade technology. The stemmed point is a projectile point made out of elongated flake or blade and with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. The earliest stemmped points were made on flakes. After blades appear at c. 27 ka, the shape of stemmped points became more standardized with one or two ridges on dorsal side and triangular cross section. The stemmed point is often called as a tanged point in western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points in Korea that have long been called 'tanged points'. Stemmed points were made from blade blanks, and became more standardized and increased in quantity over time. Flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene  [@Bae_2010; @Lee_2013; @Lee_2016]. The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; #### Park 2012 ####; @seong2015diversity; @bae2017origin].

The origin of the stemmed point is one of the most frequently discussed topics among Paleolithic research in Korea. Especially, this debate includes questions about the origin of the Late Paleolithic period in Korea, the existence and nature of Middle Paleolithic technology, and reconstructions of anatomically modern human dispersals in eastern Asia. Most Korean archaeologists had assumed that the new blade technology represented came from Shuidonggou in the northeast China because this site has one of the earliest blade industries (around 35,000 BP) in northeast Asia and it is geographically close to the Korean peninsula [seong2008tanged; @lee2013current]. However, signs of early technological change in Korea, including stemmed points and blades, have been discovered from deposits dated to 40-35 ka BP from Hwadae-ri, Hopyeong-dong, and Yongho-dong, pre-dating the Shuidonggou finds [@seong2009emergence]. The debate about origins can be summarized into two models competing to explain the appearance of the new Late Pleistocene technologies: in situ evolution [@seong2009emergence] and ‘heterogenic’ migration [@bae2010origin]. The ‘in situ’ model claims that stemmed points and other Late Paleolithic assemblages including a blade industry autonomously emerged in the south of the Korean peninsula with no external influence. Seong (2009) uses the blade-to-flake ratios of lithic assemblages to support his claim. In his view, the blade industry represents new technology while flakes indicate a continuing local one. He argues that the increased ratio of blades in lithic assemblages shows an expansion of the new technology. The migration model argues that the new blade industry, including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins. While Seong’s model (2009) proposes that the ‘heterogenic’ character is the result of indigenous development, Bae (2010) proposes that it is the result of the continuous influx of modern human populations from both Siberia and southern China. Bae’s model suggests blade technology arrived in the Korean peninsula from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP [@bae2013early]. 
<!--- I don't know how to connect this debate to our research. It is really needed? --->

@@

Our review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focussed on timing, location, origins and description of these new tools, but there has been little effort to model the land-use behaviours that generated these assemblages. We focus on the appearance of stemmed points in Korea because these are one of the earliest new technologies that definiately represents a new hunting methods because the halfting implies throwing techniques that can reach long-distance targets [@Keeley_1982; @kuhn2015artifacts]. We ask two questions to invetigate the period around the introduction of stemmed points in Korea.

First, we ask how the introduction of the new tools changed foragers' landscape use, specifically, was the use of stemmed points associated with occupation of marginal or exterme environments? The new technology might allow foragers to explore and stay in different or harsher territories by increasing mobility. For instance, Morisaki et al. argue that the transition from trapezoid to the blades and projectile points around 25 ka cal BP in north Paleo-Honshu Island enabled the foragers to extend their occupation at residential sites [@Morisaki_Izuho_Terry_Sato_2015]. In Australia during the mid-Holocene, the highly retouched and finely shaped tools combined with shafts were appeared to  broaden their occupations to offshore islands on the northeast and northwest coast [@Hiscock_1994]. Hiscock et al. insist that the appearance of mirolithcs is great example to show that people made portable and multi-functional tools to minimize the travel expense and increase tool readiness [@Hiscock_Clarkson_Mackay_2011].

Second, we focus on how the new technology changed the way people use their habitation sites. After the appearance of these new tools, did they tend to stay in one location longer, or for shorter periods, perhaps for specific purposes?. Archaeological sites indicate depositional phenomena and places where artifacts and other debris accumulated reflecting human activities in the past [@kuhn2015artifact]. The stone artifact assemblage from a site can inform us how long individuals or groups stayed and give insignts into what they did there [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface]. Stone artifacts can contain a record of raw material transport and selection and give insights into land-use behaviour [@clarkson2008lithics]. For example, stone artifact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) are often used as proxies to represent occupation patterns. Assemablges with low density but a high proportion of retouched and backed pieces would indicate the remains of 'short-term camp', while the combination of high density and low retouched pieces represent 'logistically organized basecampe' [@clark2017lithics; clark2019landscapes].

@@
